MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments

### 2022-03-08 main/MSstatsPTM_Main.tex
Small

This section summarizes the datasets in this manuscript. The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods. Two computer simulations were run which varied in experimental realism. The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies. A summary of the experiments used in this manuscript can be seen in Table \ref{tab:dataDescription}. Details of the raw data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in either MassIVE.quant or Github (with direct links available in the table).

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
&  & \multicolumn{3}{c|}{Experimental design} & Data availability \\
& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio. & (MassIVE.quant  \\
&  & & Conditions & Replicates & or R package) \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6lm}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset used in this manuscript.
`Comparison'' is the type of experimental design (group comparison, or time course).
``No. of Conditions'' states the number of conditions in the dataset.
``No. of Bio. Replicates'' states the number of biological replicates per condition.
``Data availability'' column shows the MassIVE.quant ID or github repository. 
}
\label{tab:dataDescription}
\end{table}

@@

To evaluate the proposed method in situations with biological variation and known ground truth we created synthetic datasets which mimicked a label-free experiment. Multiple simulations were generated with differing numbers of replicates and conditions per simulation. All simulations were created with a high number of modified peptide features and no missing values. This was done to see how the method performed in different experimental design scenarios. The exact numbers of replicates and conditions can be seen in Table \ref{tab:dataDescription}.

In each simulation, 1000 PTMs and 1000 unmodified proteins were generated. Each PTM and protein was created with 10 corresponding features. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein. Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.

>>>>

The benchmark datasets in this manuscript provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments. Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in either MassIVE.quant \todo{cite the paper, provide url}. Details of computer simulations are available on GitHub.

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
&  & Design & Conditions & Replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github \todo{url?}}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github \todo{url?}} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset in this manuscript.
``Data availability'' shows the ID of the MassIVE.quant repository \todo{url} or the GitHub repository. 
}
\label{tab:dataDescription}
\end{table}

@@

The simulation mimicked a label-free experiment. Synthetic datasets had a high number of modified peptide features and no missing values, while varying numbers of replicates and conditions. Specifically, each dataset had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein. Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.
